---
title: "ARIMA Forecasting on Coffee Sales"
output: html_document
---

```{r setup, include=FALSE}
library(forecast)
library(tseries)
library(ffp)
```



```{r}
dat <- read.csv("Coffe_sales.csv")
dat$Date <- as.Date(dat$Date)   


daily <- aggregate(money ~ Date, data = dat, FUN = sum)


y <- ts(daily$money, frequency = 7)
```

## 1. Fit ARIMA Model & Explain Output

```{r}
fit_arima <- auto.arima(y)
summary(fit_arima)
```

**Explanation:**

The ARIMA(2,0,2) model captures the short-term dynamics in my coffee sales data without needing difference. The second-lag auto regressive term is strong, indicating that sales from two days ago significantly influence today’s sales. 
All information criteria (AIC, AICc, BIC) support this model as a good fit. Residuals show no autocorrelation, and MASE < 1 confirms that the model performs better than a naïve benchmark. Overall, the ARIMA model is appropriate and provides reliable forecasts for my dataset.

## 2. Residual Analysis

```{r}
res_arima <- residuals(fit_arima)
res_arima <- na.omit(res_arima)


plot(res_arima, main="Residuals (ARIMA)")
hist(res_arima, main="Histogram (ARIMA Residuals)")
acf(res_arima, main="ACF (ARIMA Residuals)")
```

## 3. Forecast Next 5 Periods

```{r}
arima_forecast <- forecast(fit_arima, h=5)
arima_forecast
plot(arima_forecast, main="ARIMA Forecast for Next 5 Periods")
```

## 4. Accuracy of ARIMA Model

```{r}
accuracy(fit_arima)
```
